---
title: "Biomarker_Cleaning"
output:
  html_document:
    toc: true
    theme: united
date: "2023-06-28"
---

This file reads in the merged biomarker file and cleans the data for modelling. This file also cleans the viral load data.

Read in:

  * m4_dem.csv
  * m4_fin.csv
  * f3109.sas7bdat (viral load)
  * hxw0101.sas7bdat

Output: 

  * ivs_stand_flipped.csv (Participant and Biomarker Data ready for modelling)

  * ivs_stand_flipped.csv (csv file of biomarkers and participant info ready for modelling)


# Libraries
```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
# library(MASS)
library("skimr")
library("dplyr")
library("RcmdrMisc")
library("haven")
library("sjPlot")
library(ggplot2)
library(purrr)
library(tidyr)
library(ggrepel)
library(corrplot)
library(dplyr)
```

# Import Data 
```{r}
m4_dem <- read.csv(paste0(root, "/Exported_Data/m4_dem.csv"))
m4_fin <- read.csv(paste0(root, "/Exported_Data/m4_fin.csv"))

v_load <- read_sas(paste0(root, "/data/f3109.sas7bdat"))
hxw <- read_sas(paste0(root, "/data/hxw0101.sas7bdat"))

```


# Merge Biomarkers and Demographic
```{r}
bm_dem <- merge(m4_fin, m4_dem, by = c("publicID", "week"))

skim(bm_dem)
```

# Check Viral Load and Maximum Viral load
```{r}

skim(hxw)
skim(hxw_old)

```
There are duplicates in the v_load data. Let's find them and get only the most recent of the duplicates



```{r}
hist(v_load$week)
table(v_load$week)
skim(v_load)
skim(v_load_old)
```


# Find duplicate in HIV Viral Load data
```{r}
# Get duplicates
duplicates <- v_load %>%
  group_by(publicID, week) %>%
  filter(duplicated(publicID, week) | duplicated(publicID, week, fromLast = TRUE))

# Ungroup the duplicates data frame
ungrouped_duplicates <- duplicates %>% 
  ungroup()

# Filter ungrouped duplicates to select complete rows
complete_rows_duplicates <- ungrouped_duplicates %>%
  filter(complete.cases(.))

# Select the row with the higher value of specage for each duplicate
selected_rows <- complete_rows_duplicates %>%
  group_by(publicID, week) %>%
  slice(which.max(specage))

selected_rows
```

# Anti-Join and Re-Join HIV Viral Load
```{r}
# Perform anti-join
anti_1 <- anti_join(v_load, duplicates, by = c("publicID", "week"))
rejoin_1 <- rbind(anti_1, selected_rows)

# Check to see that all duplicates are gone
rejoin_1 %>%
  group_by(publicID, week) %>%
  filter(duplicated(publicID, week) | duplicated(publicID, week, fromLast = TRUE))
```
rejoin_1 is the no-duplicates HIV viral load dataframe.


# Merge Viral Load and Maximum Viral load
```{r}
vl_pres_past <- merge(rejoin_1, hxw, by = c("publicID"))

```
This data frame has both current viral load and past peak viral load and CD4 counts. 

# Merge bm_dem and vl_pres_past
```{r}
vl_pres_past$week <- vl_pres_past$week.x
ivs <- merge(bm_dem, vl_pres_past, by = c("publicID", "week"))

skim(ivs)
```

# Check and Remove Duplicates
Take the most updated Marker
```{r}
ivs %>%
  group_by(publicID, week) %>%
  filter(n() > 1)

```

## See Trajectories of each biomarker
```{r}
# # Split between male and female for standardization
# ivs_m <- ivs %>% filter(sex == "M") 
# ivs_f <- ivs %>% filter(sex == "F") 
# 
# # Standardize the columns within the specified range
# ivs_stand_m <- ivs_m %>%
#   mutate(across(syst_avg:mwaist_avg, scale))
# 
# ivs_stand_f <- ivs_f %>%
#   mutate(across(syst_avg:mwaist_avg, scale))
# 
# ivs_stand <- rbind(ivs_stand_m, ivs_stand_f)
# 
# # Make Weeks a factor
# ivs_stand$week_fac <- as.factor(ivs_stand$week)
# 
# # New data frame with averages
# avg_ivs_stand <- ivs_stand %>%
#   group_by(week) %>%
#   summarize(across(syst_avg:mwaist_avg, mean, na.rm = TRUE))
# 
# # Reshape the data into long format
# avg_ivs_stand_long <- avg_ivs_stand %>% gather(biomarker, value, syst_avg:mwaist_avg)
# 
# # Choose the time point for labeling (e.g., week 96)
# label_time <- 96
# 
# # Filter the data to include only the chosen time point
# avg_ivs_stand_long_labels <- avg_ivs_stand_long %>% filter(week == label_time)
# 
# # Plot the overlayed biomarker values over time with direct labels
# ggplot(avg_ivs_stand_long, aes(x = week, y = value, color = biomarker)) +
#   geom_line() +
#   xlab("Week") +
#   ylab("Biomarker Value") +
#   ggtitle("Overlayed Biomarker Values over Time")

```

# Graph only biomarkers with a negative slope
```{r}
# avg_ivs_stand
# 
# #Any negative values here have a negative slope (gets lower with age)
# avg_ivs_stand %>% 
#   filter(week == 0 | week == 96) %>%
#   summarize(across(syst_avg:mwaist_avg, ~ .[week == 96] - .[week == 0]))
# 
# # Reshape the data into long format
# 
# neg_cols <- avg_ivs_stand %>% select(week, sgotval, bunval, ckval, ftcval, ftgval, fldlval, fhdlval, 
#                                      hemocv, mcv, wbccv, lymph, mono, eosin, baso, platcv, alympcnt) 
# 
# # Reshape the data into long format
# avg_ivs_stand_long_neg <- neg_cols %>% gather(biomarker, value, sgotval:alympcnt)
# 
# # Plot the overlayed biomarker values over time with direct labels
# ggplot(avg_ivs_stand_long_neg, aes(x = week, y = value, color = biomarker)) +
#   geom_line() +
#   xlab("Week") +
#   ylab("Biomarker Value") +
#   ggtitle("Overlayed Biomarker Values over Time")

```

# Find Correlation between each biomarker and age. 
Take out any that are very weakly correlated (0.1 or less)
```{r}
colnames(ivs)
ivs <- ivs %>%
  mutate(across(c(specage.y, syst_avg:mwaist_avg), as.numeric))
# Select the specific column and range of columns
selected_columns <- ivs %>%
  select(specage.y, syst_avg:mwaist_avg, -study)

# Calculate the correlation matrix
cor_matrix <- cor(selected_columns,method = "spearman", use = "pairwise.complete.obs")
cor_table <- as.data.frame(as.table(cor_matrix))

cor_table

cor.test(ivs$anccv, ivs$wbccv, method = "spearman", use = "pairwise.complete.obs")
ivs %>% select(anccv, wbccv) %>% filter(complete.cases(.))


cor_table_age <- cor_table %>% 
  filter(Var1 == "specage.y") %>% 
  arrange(abs(Freq))

median(abs(cor_table_age$Freq))

# Subset the top half of correlation
subset(cor_table_age, abs(Freq) <= median(abs(cor_table_age$Freq)))

# Get list of not very correlated biomarkers
bms_not_cor_list <- subset(cor_table_age, abs(Freq) <= median(abs(cor_table_age$Freq))) %>%
  pull(Var2)

```


## Correlation matrix of Inflammation and Infections BM's
```{r}
# cor_matrix <- cor(ivs %>% select(lipval,rbc,mcv, anccv, lymph, platcv, hema, wbccv), 
#                   method = "spearman", use = "pairwise.complete.obs")

cor_matrix <- cor(ivs %>% select(syst_avg:mwaist_avg, -study), 
                  method = "spearman", use = "pairwise.complete.obs")

# Define your own list of strings for variable names
# new_variable_names <- c("Lipase Value", "Red Blood Cell Count", "Mean Corpuscular Volume", "Neutrophil Value", "Lymphocyte Value", "Platelet Value", "Hematocrit Value", "White Blood Cell Count")

# Rename the column names of the correlation matrix
# colnames(cor_matrix) <- new_variable_names
# rownames(cor_matrix) <- new_variable_names

# Define the row names and groups
row_names <- rownames(cor_matrix)
groups <- c("Included", "Excluded")  # Specify your groups here

# Create a custom color palette for the groups
group_colors <- c("Group1" = "red", "Group2" = "blue")  # Define colors for each group

# Map row names to corresponding colors
row_colors <- group_colors[groups]

# Use corrplot to create a colored correlation matrix with renamed variables
corrplot(cor_matrix, method = "color", type = "upper", order = "hclust",
         tl.col = "black", tl.srt = 45, tl.cex = 0.7,  # Adjust label size (0.7)
         addCoef.col = "black", 
         number.cex = 0.3,
         diag = FALSE,
         width = 0.6)  # Adjust cell width (0.6)
 
```


## Graph Correlation by Chronological Age at Baseline
This deals with independence of correlation assumption
```{r}
cor_assump_cols <- ivs %>% 
  filter(week == 0) %>%
  select(specage.y, syst_avg:mwaist_avg)

# Calculate the correlation matrix
cor_matrix <- cor(cor_assump_cols,method = "spearman", use = "pairwise.complete.obs") 
cor_table <- as.data.frame(as.table(cor_matrix))

subset(cor_table, abs(Freq) <= 0.07) %>% arrange(Freq) %>% filter(grepl("specage.y", Var1))


```
From baseline, BM's that have a correlation with CA less than 0.07:

  * baso
  * ckval
  * bunval
  * wbccv
  * mono 
  * sgptval


Take out 

  * baso
  * mono
  * wbccv



# Correlations of biomarkers
```{r}
ivs <- ivs %>% mutate_at(vars(syst_avg:bmi), as.double)

# from correlation with chronological age
ivs_removed <- ivs %>% select(-c(mwaist_avg, weight_avg, height_avg, baso, mono, wbccv))

cor_matrix <- cor(ivs_removed %>% select(syst_avg:bmi), method = "spearman", use = "pairwise.complete.obs")

cor_table <- as.data.frame(as.table(cor_matrix))
subset(cor_table, abs(Freq) > 0.8) %>% arrange(Freq)

cor.test(ivs$hema, ivs$hemocv, method = "spearman", use = "pairwise.complete.obs")
cor.test(ivs$fldlval, ivs$ftcval, method = "spearman", use = "pairwise.complete.obs")

```
In choosing which one to take out, look at the trajectories... If it is more linear, then keep that one



## Find correlation at baseline 
This deals with independence of correlation assumption
```{r}
cor_assump_cols_bm_only <- cor_assump_cols %>% mutate_at(vars(syst_avg:mwaist_avg), as.double)
cor_assump_cols_bm_only <- cor_assump_cols_bm_only %>% select(-specage.y)

cor_matrix <- cor(cor_assump_cols_bm_only %>% select(syst_avg:mwaist_avg), 
                  method = "spearman", use = "pairwise.complete.obs")

cor_table <- as.data.frame(as.table(cor_matrix))
subset(cor_table, abs(Freq) > 0.8) %>% arrange(Freq)

```


Take out:

  * lymph
  * anccv
  * hema
  * height
  * weight


# Final List of Biomarkers

Take out:  

  * Low correlation with age
    + baso
    + mono
    + wbccv
  * High correlation with other BM
    + lymph
    + anccv
    + hema
    + height
    + weight
    
# Remove Outliers in Baseline
There are a lot of extreme upper values. Get rid of them in baseline since we will standardize based on baseline. 
```{r}
ivs_bm_names <- ivs %>% dplyr::select(c(syst_avg:mwaist_avg, -alympcnt)) %>% names()

for (bm in ivs_bm_names) {
  
  hist(ivs %>% select(bm))
}


# Step 1: Scale the values within each column 
scaled_ivs <- ivs %>%
  mutate(across(c(syst_avg:bmi, -alympcnt), scale))

# Step 2: Replace values greater than the absolute value of 4 with NA
scaled_ivs <- scaled_ivs %>%
  mutate(across(c(syst_avg:bmi), ~ ifelse(abs(.) > 4, NA, .)))

for (bm in ivs_bm_names) {
  
  hist(scaled_ivs %>% select(bm))
}


```


# Backwards Selection 
## V3
```{r}
scaled_ivs

scaled_ivs_comp <- scaled_ivs %>% 
  dplyr::select(syst_avg:platcv, weight_avg:mwaist_avg, specage.y, -study) 

scaled_ivs_comp <- scaled_ivs_comp %>%
  mutate_all(function(x) if_else(is.na(x), mean(x, na.rm = TRUE), x))

scaled_ivs_comp


# Backward Selection using step function with AIC
# Perform backward selection
model <- lm(formula = specage.y ~ ., 
              data = scaled_ivs_comp) 

# Assuming 'model' is already defined as you mentioned
# Specify the formula and data for the initial model
initial_formula <- formula(model)
initial_formula

library(MASS)
# Perform stepwise selection with AIC in both directions
final_model <- stepAIC(model, direction = "backward")

# Print the final selected model
summary(final_model)


# Get Final List
stepaic_list <- all.vars(final_formula)[-1]
scaled_ivs_comp %>% names()
stepaic_list

setdiff(scaled_ivs_comp %>% names(), stepaic_list)

detach("package:MASS", unload= TRUE)
```


# Take out selected BM's and standardize
Standardization will be based on the baseline value.
```{r}
ivs$map <- (2 * ivs$dia_avg + ivs$syst_avg) / 3

ivs<- ivs %>%
  relocate(map, .before = mwaist_avg)

ivs_removed <- ivs %>% select(-c(baso, mono, wbccv, lymph, anccv, hema, 
                                             height_avg, weight_avg, dia_avg, syst_avg))

list_v2_remove <- union(c("baso", "mono", "wbccv", "lymph", "anccv", "hema", "height_avg", 
                          "weight_avg", "dia_avg", "syst_avg"),
                        bms_not_cor_list)


ivs_removed_v2 <- ivs %>% select(-list_v2_remove)

# Step AIC Selected
ivs_bm_all_names <- ivs %>% select(syst_avg:alympcnt, weight_avg:mwaist_avg) %>% names()

ivs_removed_v3 <- ivs %>% 
  select(-c(setdiff(ivs_bm_all_names,stepaic_list)))

```


Final List of Biomarkers:

V3:
Metabolic Markers:
- Fasting Triglycerides Value
- Fasting HDL Cholesterol Value

Markers of Inflammation and Infection:
- Absolute Neutrophil Count
- Eosinphil Percentage
- Absolute Platelets Value

Markers of Organ Function
- Alanine Aminotransferase (ALT/SGPT) Value
- Aspartate Aminotransferase (AST/SGOT) Value
- Blood Urea Nitrogen (BUN) (mg/dl)
- Serum Creatinine Kinase 
- Creatine Kinase Value

Cardiovascular Markers
- Systolic Average
- Diastolic Average
- MAP

Anthropometric Markers
- BMI
- Weight
- Height

BM's Taken out:
- lipval
- fldlval
- hemocv
- rbc
- mcv
- ftcval








## Scale Biomarkers Function
```{r}
scale_biomarkers <- function(data, bms) {
    # Split between male and female for standardization
  # Standardization 
  ivs_m_v2 <- data %>% filter(sex == "M") 
  ivs_f_v2 <- data %>% filter(sex == "F") 
  
  # Columns to standardize (excluding the 'week' column)
  biomarker_cols_v2 <- ivs_m_v2 %>% select(bms) %>% names()
  
  ivs_m_0_v2 <- ivs_m_v2 %>% filter(week == 0)
  ivs_f_0_v2 <- ivs_f_v2 %>% filter(week == 0)
  # Extract the mean and standard deviation for each column at week 0
  ivs_m_0_summary_v2 <- ivs_m_0_v2 %>%
    summarize(across(bms, list(mean = ~ mean(., na.rm = TRUE), 
                                               sd = ~ sd(., na.rm = TRUE))))
  ivs_f_0_summary_v2 <- ivs_f_0_v2 %>%
    summarize(across(bms, list(mean = ~ mean(., na.rm = TRUE), 
                                               sd = ~ sd(., na.rm = TRUE))))
  
  
  scale_column <- function(x, mean_val, sd_val) {
    (x - mean_val) / sd_val
  }
  
  ivs_m_scale_v2 <- ivs_m_v2
  ivs_f_scale_v2 <- ivs_f_v2
  
  
  # For loop to scale each column based on mean and sd at baseline
  for (bm in bms) {
    
    # Males
    mean_val_m_v2 <- ivs_m_0_summary_v2 %>%
      select(matches(paste0(bm, "_mean")))
    sd_val_m_v2 <- ivs_m_0_summary_v2 %>%
      select(matches(paste0(bm, "_sd")))
    
    # Females
    mean_val_f_v2 <- ivs_f_0_summary_v2 %>%
      select(matches(paste0(bm, "_mean")))
    sd_val_f_v2 <- ivs_f_0_summary_v2 %>%
      select(matches(paste0(bm, "_sd")))
    
    # Scale the selected columns using mutate_at
    ivs_m_scale_v2 <- ivs_m_scale_v2 %>%
      mutate_at(vars(all_of(bm)), ~ scale_column(., mean_val_m_v2[1,1], sd_val_m_v2[1,1]))
    ivs_f_scale_v2 <- ivs_f_scale_v2 %>%
      mutate_at(vars(all_of(bm)), ~ scale_column(., mean_val_f_v2[1,1], sd_val_f_v2[1,1]))
  
  }
  
  # Final scaled dataframe
  ivs_stand_v2 <- rbind(ivs_m_scale_v2, ivs_f_scale_v2)
  
  return(ivs_stand_v2)
}


```



## Standardize V1
```{r}

# Split between male and female for standardization
# Standardization 
ivs_m <- ivs_removed %>% filter(sex == "M") 
ivs_f <- ivs_removed %>% filter(sex == "F") 

# Columns to standardize (excluding the 'week' column)
biomarker_cols <- ivs_m %>% select(sgptval:mwaist_avg) %>% names()

ivs_m_0 <- ivs_m %>% filter(week == 0)
ivs_f_0 <- ivs_f %>% filter(week == 0)

# Extract the mean and standard deviation for each column at week 0
ivs_m_0_summary <- ivs_m_0 %>%
  summarize(across(sgptval:mwaist_avg, list(mean = ~ mean(., na.rm = TRUE), 
                                             sd = ~ sd(., na.rm = TRUE))))
ivs_f_0_summary <- ivs_f_0 %>%
  summarize(across(sgptval:mwaist_avg, list(mean = ~ mean(., na.rm = TRUE), 
                                             sd = ~ sd(., na.rm = TRUE))))


scale_column <- function(x, mean_val, sd_val) {
  (x - mean_val) / sd_val
}

ivs_m_scale <- ivs_m
ivs_f_scale <- ivs_f

# For loop to scale each column based on mean and sd at baseline
for (bm in biomarker_cols) {
  
  # Males
  mean_val_m <- ivs_m_0_summary %>%
    select(matches(paste0(bm, "_mean")))
  sd_val_m <- ivs_m_0_summary %>%
    select(matches(paste0(bm, "_sd")))
  
  # Females
  mean_val_f <- ivs_f_0_summary %>%
    select(matches(paste0(bm, "_mean")))
  sd_val_f <- ivs_f_0_summary %>%
    select(matches(paste0(bm, "_sd")))
  
  # Scale the selected columns using mutate_at
  ivs_m_scale <- ivs_m_scale %>%
    mutate_at(vars(all_of(bm)), ~ scale_column(., mean_val_m[1,1], sd_val_m[1,1]))
  ivs_f_scale <- ivs_f_scale %>%
    mutate_at(vars(all_of(bm)), ~ scale_column(., mean_val_f[1,1], sd_val_f[1,1]))

}

# Final scaled dataframe
ivs_stand <- rbind(ivs_m_scale, ivs_f_scale)

```

## Standardize V2
```{r}

# Split between male and female for standardization
# Standardization 
ivs_m_v2 <- ivs_removed_v2 %>% filter(sex == "M") 
ivs_f_v2 <- ivs_removed_v2 %>% filter(sex == "F") 

# Columns to standardize (excluding the 'week' column)
biomarker_cols_v2 <- ivs_m_v2 %>% select(cretval:platcv, bmi:mwaist_avg) %>% names()

ivs_m_0_v2 <- ivs_m_v2 %>% filter(week == 0)
ivs_f_0_v2 <- ivs_f_v2 %>% filter(week == 0)
# Extract the mean and standard deviation for each column at week 0
ivs_m_0_summary_v2 <- ivs_m_0_v2 %>%
  summarize(across(c(cretval:platcv, bmi:mwaist_avg), list(mean = ~ mean(., na.rm = TRUE), 
                                             sd = ~ sd(., na.rm = TRUE))))
ivs_f_0_summary_v2 <- ivs_f_0_v2 %>%
  summarize(across(c(cretval:platcv, bmi:mwaist_avg), list(mean = ~ mean(., na.rm = TRUE), 
                                             sd = ~ sd(., na.rm = TRUE))))


scale_column <- function(x, mean_val, sd_val) {
  (x - mean_val) / sd_val
}

ivs_m_scale_v2 <- ivs_m_v2
ivs_f_scale_v2 <- ivs_f_v2


# For loop to scale each column based on mean and sd at baseline
for (bm in biomarker_cols_v2) {
  
  # Males
  mean_val_m_v2 <- ivs_m_0_summary_v2 %>%
    select(matches(paste0(bm, "_mean")))
  sd_val_m_v2 <- ivs_m_0_summary_v2 %>%
    select(matches(paste0(bm, "_sd")))
  
  # Females
  mean_val_f_v2 <- ivs_f_0_summary_v2 %>%
    select(matches(paste0(bm, "_mean")))
  sd_val_f_v2 <- ivs_f_0_summary_v2 %>%
    select(matches(paste0(bm, "_sd")))
  
  # Scale the selected columns using mutate_at
  ivs_m_scale_v2 <- ivs_m_scale_v2 %>%
    mutate_at(vars(all_of(bm)), ~ scale_column(., mean_val_m_v2[1,1], sd_val_m_v2[1,1]))
  ivs_f_scale_v2 <- ivs_f_scale_v2 %>%
    mutate_at(vars(all_of(bm)), ~ scale_column(., mean_val_f_v2[1,1], sd_val_f_v2[1,1]))

}

# Final scaled dataframe
ivs_stand_v2 <- rbind(ivs_m_scale_v2, ivs_f_scale_v2)

```


## V3
```{r}
ivs_stand_v3 <- scale_biomarkers(ivs_removed_v3, ivs_removed_v3 %>% select(sgptval:mwaist_avg) %>% names())
```


# Cleaning
## V1
```{r}
# Make Weeks a factor
ivs_stand$week_fac <- as.factor(ivs_stand$week)

# Make BM's numeric
ivs_stand <- ivs_stand %>%
  mutate_at(vars(sgptval:mwaist_avg), as.numeric)

# Clean out some unnecessary columns
ivs_stand <- ivs_stand %>% select(-c(X.x, X.y, study))

colnames(ivs_stand)

```
  * Metabolic Markers
    + ftcval (Fasting total cholesterol val mg/dl)
    + ftgval (Fasting triglycerides value mg/dL)
    + fldlval (Fasting LDL value mg/dL)
    + fhdlval (Fasting HDL value mg/dL)
  * Markers of Inflammation and Infection
    + hemocv (Hemoglobin value (g/dl))
    + rbc (Red Blood Cell count)
    + anccv (Abs count neutrophil value (mm^3))
    + platcv (Platelets value (mm^3))
    + alympcnt (Abs Lymphocyte count)
    + lipval (Lipase)
    + mcv (mean corpuscular volume)
    + neutcv (neutrophil value)
    + eosin (eosinphil value)
  * Markers of Organ Function
    + sgotval (Aspartate Aminotransferase)
    + sgptval (Alanine Aminotransferase)
    + cretval (Serum Creatinine Kinase)
    + bunval (Blood Urea Nitrogen)
    + ckval (Creatine Kinase)
  * Cardiovascular Markers
    + MAP (from syst_avg and dia_avg)
  * Anthropometric Markers
    + BMI
    + mwaist_avg
  
21 biomarkers in total


## V2
```{r}
# Make Weeks a factor
ivs_stand_v2$week_fac <- as.factor(ivs_stand_v2$week)

# Make BM's numeric
ivs_stand_v2 <- ivs_stand_v2 %>%
  mutate_at(vars(cretval:platcv, bmi:mwaist_avg), as.numeric)


colnames(ivs_stand_v2)
```


## V3
```{r}
# Make Weeks a factor
ivs_stand_v3$week_fac <- as.factor(ivs_stand_v3$week)

# Make BM's numeric
ivs_stand_v3 <- ivs_stand_v3 %>%
  mutate_at(ivs_removed_v3 %>% select(sgptval:mwaist_avg) %>% names(), as.numeric)

colnames(ivs_stand_v3)
```


# Remove Outliers in Baseline
## V1
There are a lot of extreme upper values. Get rid of them in baseline since we will standardize based on baseline. 
```{r}
ivs_stand %>%
  mutate(across(sgptval:mwaist_avg, ~ ifelse(abs(.) > 4, NA, .))) %>%
  summarize(total_na = sum(is.na(across(everything()))))

ivs_stand %>%
  summarize(total_na = sum(is.na(across(everything()))))

ivs_stand_removed <- ivs_stand %>%
  mutate(across(sgptval:mwaist_avg, ~ ifelse(abs(.) > 4, NA, .)))

ivs_bm_names <- ivs_stand_removed %>% select(sgptval:mwaist_avg, -specage.x) %>% names()

for (bm in ivs_bm_names) {
  
  hist(ivs_stand_removed %>% select(bm))
}


```

## V2
There are a lot of extreme upper values. Get rid of them in baseline since we will standardize based on baseline. 
```{r}

ivs_stand_v2 %>%
  mutate(across(cretval:mwaist_avg, ~ ifelse(abs(.) > 4, NA, .))) %>%
  summarize(total_na = sum(is.na(across(everything()))))

ivs_stand_v2 %>%
  summarize(total_na = sum(is.na(across(everything()))))

ivs_stand_removed_v2 <- ivs_stand_v2 %>%
  mutate(across(cretval:mwaist_avg, ~ ifelse(abs(.) > 4, NA, .)))

ivs_bm_names_v2 <- ivs_stand_removed_v2 %>% select(cretval:mwaist_avg) %>% names()

for (bm in ivs_bm_names_v2) {
  
  hist(ivs_stand_removed_v2 %>% select(bm))
}


```

## V3
There are a lot of extreme upper values. Get rid of them in baseline since we will standardize based on baseline. 
```{r}

ivs_stand_v3 %>%
  mutate(across(sgptval:mwaist_avg, ~ ifelse(abs(.) > 4, NA, .))) %>%
  summarize(total_na = sum(is.na(across(everything()))))

ivs_stand_v3 %>%
  summarize(total_na = sum(is.na(across(everything()))))

ivs_stand_removed_v3 <- ivs_stand_v3 %>%
  mutate(across(sgptval:mwaist_avg, ~ ifelse(abs(.) > 4, NA, .)))

ivs_bm_names_v3 <- ivs_stand_removed_v3 %>% select(sgptval:mwaist_avg) %>% names()

for (bm in ivs_bm_names_v3) {
  
  hist(ivs_stand_removed_v3 %>% select(bm))
}


```



# Directionality
## V1
```{r}
# New data frame with averages
# avg_ivs_stand <- ivs_stand %>%
#   group_by(week) %>%
#   summarize(across(syst_avg:mwaist_avg, mean, na.rm = TRUE))
# 
# # Reshape the data into long format
# avg_ivs_stand_long <- avg_ivs_stand %>% gather(biomarker, value, syst_avg:bmi)
# 
# # Choose the time point for labeling (e.g., week 96)
# label_time <- 96
# 
# # Filter the data to include only the chosen time point
# avg_ivs_stand_long_labels <- avg_ivs_stand_long %>% filter(week == label_time)
# 
# # Plot the overlayed biomarker values over time with direct labels
# ggplot(avg_ivs_stand_long, aes(x = week, y = value, color = biomarker)) +
#   geom_line() +
#   xlab("Week") +
#   ylab("Biomarker Value") +
#   ggtitle("Overlayed Biomarker Values over Time")

biomarkers <- ivs_stand_removed %>%
  select(sgptval:mwaist_avg) %>%
  names()


for (bm in biomarkers) {
  # Plot the overlayed biomarker values over time with direct labels as dots
  print(ggplot(ivs_stand_removed %>% filter(week == 0), aes_string(x = "specage.y", y = bm, color = "sex")) +
    geom_point() +  # Plot data points as dots
    geom_smooth(method = "lm", se = FALSE) +  # Add a regression line
    xlab("Age") +
    ylab("Biomarker Value") +
    ggtitle(paste("Overlayed Biomarker Values over Time -", bm)))
}


```

Markers that need to be flipped:
+ sgotval (Higher value shows liver damage)
+ fldlval (Higher is worse)
+ platcv (Higher may be because of infection)

## V2
```{r}

biomarkers_v2 <- ivs_stand_removed_v2 %>%
  select(cretval:mwaist_avg) %>%
  names()


for (bm in biomarkers_v2) {
  # Plot the overlayed biomarker values over time with direct labels as dots
  print(ggplot(ivs_stand_removed_v2 %>% filter(week == 0), aes_string(x = "specage.y", y = bm, color = "sex")) +
    geom_point() +  # Plot data points as dots
    geom_smooth(method = "lm", se = FALSE) +  # Add a regression line
    xlab("Age") +
    ylab("Biomarker Value") +
    ggtitle(paste("Overlayed Biomarker Values over Time -", bm)))
}
```


## V3
```{r}

biomarkers_v3 <- ivs_stand_removed_v3 %>%
  select(sgptval:mwaist_avg) %>%
  names()


for (bm in biomarkers_v3) {
  # Plot the overlayed biomarker values over time with direct labels as dots
  print(ggplot(ivs_stand_removed_v3 %>% filter(week == 0), aes_string(x = "specage.y", y = bm, color = "sex")) +
    geom_point() +  # Plot data points as dots
    geom_smooth(method = "lm", se = FALSE) +  # Add a regression line
    xlab("Age") +
    ylab("Biomarker Value") +
    ggtitle(paste("Overlayed Biomarker Values over Time -", bm)))
}
```


# Flip negative sloped biomarkers and make scaled CA column
## V1
```{r}
# Multiply selected columns by -1
ivs_stand_flipped <- ivs_stand_removed %>%
  mutate_at(vars(ftcval, fldlval, platcv), ~ . * -1)   

# Create scaled choronological age column
ivs_stand_flipped$scaled_age <- scale(ivs_stand_flipped$specage.y)

```

## V2
```{r}
# Multiply selected columns by -1
ivs_stand_flipped_v2 <- ivs_stand_removed_v2 %>%
  mutate_at(vars(fhdlval, platcv), ~ . * -1)   

# Create scaled choronological age column
ivs_stand_flipped_v2$scaled_age <- scale(ivs_stand_flipped_v2$specage.y)
```

## V3
```{r}
# Multiply selected columns by -1
ivs_stand_flipped_v3 <- ivs_stand_removed_v3 %>%
  mutate_at(vars(fhdlval, sgotval, platcv), ~ . * -1)   

# Create scaled choronological age column
ivs_stand_flipped_v3$scaled_age <- scale(ivs_stand_flipped_v3$specage.y)
```

## Check by participant (male)
### V1
```{r}

for (bm in biomarkers) {
  
  print(ggplot(ivs_stand_flipped %>% filter(sex == "F"), aes_string(x = "specage.y", y = bm, 
                                                                    group = "publicID")) +
  geom_point() +
  geom_line() +
  xlab("Age") +
  ylab(bm) +
  ggtitle("Scatterplot with Lines for Timepoints"))
}

```

Biomarkers that seem to decrease significantly with age (both m and f):

  * fldlval

Too straight...

  * sgotval
  * bunval
  * ckval
  * cretval
  * ftgval
  * eosin
  
## V3
```{r}
for (bm in biomarkers_v3) {
  
  print(ggplot(ivs_stand_flipped_v3 %>% filter(sex == "F"), aes_string(x = "specage.y", y = bm, 
                                                                    group = "publicID")) +
  geom_point() +
  geom_line() +
  xlab("Age") +
  ylab(bm) +
  ggtitle("Scatterplot with Lines for Timepoints"))
}
```


### V3 Export Removed BM's
```{r}
write.csv(ivs_stand_flipped, paste0(root, "Exported_Data/ivs_stand_flipped.csv"))



biomarkers_V3_removed <- scaled_ivs_comp %>% 
  select(lipval, fldlval, hemocv, hema, rbc, mcv, ftcval, mono, baso, eosin, lymph, neutcv) %>% 
  names()

# Specify the list of titles
titles <- c("Lipase", "Fasting LDL Cholesterol (mg/dL)", "Hemoglobin (g/dl)", "Hematocrit", 
            "Red Blood Cell Count", " Mean Corpuscular Value (Microns)", 
            "Fasting Total Cholesterol (mg/dL)", "Monocytes Value", "Basophil Value", 
            "Eosinphil Value", "Lymphocyte Value", "Neutrophil Percentage")  # Replace with your titles

for (i in seq_along(biomarkers_V3_removed)) {
  bm <- biomarkers_V3_removed[i]
  title <- paste("Scatterplot by Sex for", titles[i])

  p <- ggplot(scaled_ivs, aes(x = specage.y, group = publicID)) +
    geom_point(aes_string(y = bm, color = "sex"), size = 1) +
    geom_line(aes_string(y = bm, color = "sex"), size = 0.5) +
    facet_grid(sex ~ .) +
    xlab("Age") +
    ylab(bm) +
    ggtitle(title) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 16),  # Increase title font size
      axis.text.x = element_text(size = 14),  # Increase x-axis label font size
      axis.text.y = element_text(size = 14),  # Increase y-axis label font size
      axis.title.x = element_text(size = 16),  # Increase x-axis title font size
      axis.title.y = element_text(size = 16)   # Increase y-axis title font size
    )

  print(p)
}




```
  
### V3 Export Kept BM's
```{r}
write.csv(ivs_stand_flipped, paste0(root, "Exported_Data/ivs_stand_flipped.csv"))


biomarkers_V3 <- scaled_ivs_comp %>% 
  dplyr::select(-c(lipval, fldlval, hemocv, hema, rbc, mcv, ftcval, mono, baso, eosin, lymph, 
                   neutcv, weight_avg, height_avg)) %>% 
  names()

# Specify the list of titles
titles <- c("Systolic BP", "Diastolic BP", "ALT/SGPT", "AST/SGOT", 
            "Blood Urea Nitrogen", " Creatine Kinase", 
            "Serum Creatinine", "Fasting Triglycerides (mg/dL)", "Fasting HDL Value (mg/dL)", 
            "White Blood Cell Count", "Neutrophil Count", "Platelet Value", "BMI", 
            "Waist Circumference")  # Replace with your titles

for (i in seq_along(biomarkers_V3)) {
  bm <- biomarkers_V3[i]
  title <- paste("Scatterplot by Sex for", titles[i])

  p <- ggplot(scaled_ivs, aes(x = specage.y, group = publicID)) +
    geom_point(aes_string(y = bm, color = "sex"), size = 1) +
    geom_line(aes_string(y = bm, color = "sex"), size = 0.5) +
    facet_grid(sex ~ .) +
    xlab("Age") +
    ylab(bm) +
    ggtitle(title) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 16),  # Increase title font size
      axis.text.x = element_text(size = 14),  # Increase x-axis label font size
      axis.text.y = element_text(size = 14),  # Increase y-axis label font size
      axis.title.x = element_text(size = 16),  # Increase x-axis title font size
      axis.title.y = element_text(size = 16)   # Increase y-axis title font size
    )

  print(p)
}




```


# Scale Starting Age and week
## V1
```{r}
ivs_stand_flipped$scale_base_age <- ivs_stand_flipped$entryage - 7
ivs_stand_flipped$week_scale <- ivs_stand_flipped$week/48
ivs_stand_flipped
```

## V2
```{r}
ivs_stand_flipped_v2$scale_base_age <- ivs_stand_flipped_v2$entryage - 7
ivs_stand_flipped_v2$week_scale <- ivs_stand_flipped_v2$week/48
```

## V3
```{r}
ivs_stand_flipped_v3$scale_base_age <- ivs_stand_flipped_v3$entryage - 7
ivs_stand_flipped_v3$week_scale <- ivs_stand_flipped_v3$week/48
```

# Add SES back into
# V1
```{r}
ivs_stand_flipped
```





# Export
```{r}
# ivs_stand_flipped <- ivs_stand_flipped %>% select(-c(sgotval, bunval, ckval, cretval, ftgval, eosin))

write.csv(ivs_stand_flipped, "/Users/hansoochang/Drexel/PHACS:AMP/data/ivs_stand_flipped.csv")
write.csv(ivs_stand_flipped_v2, "/Users/hansoochang/Drexel/PHACS:AMP/data/ivs_stand_flipped_v2.csv")
write.csv(ivs_stand_flipped_v3, "/Users/hansoochang/Drexel/PHACS:AMP/data/ivs_stand_flipped_v3.csv")

# Assuming 'participant_id' is the unique identifier for each participant
ivs_stand_flipped_v3 %>%
  group_by(publicID) %>%
  summarize(num_week = n_distinct(week)) %>%
  filter(num_week == 3) 

```




