---
title: "Initial Biomarker"
output:
  html_document:
    toc: true
    theme: united
date: "2023-01-30"
---

This file merges all the necessary biomarker files. Takes care of NaN values and unnecssary columns

Read in:
evw0173.sas7bdat (blood pressure)
bw0070.sas7bdat (Routine Chemistries)
lbw0073.sas7bdat (Routine Hematologies)
evw0169.sas7bdat (Height, Weight, Body measurements)
dmw0080.sas7bdat (HIV Status)
master.sas7bdat (Master Sheet for Participant Demographics)

Output:
m4_dem.csv (Participant demographics and all merged biomarkers)
m4_fin.csv (Merged biomarkers only)


# Libraries and wd
```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
library("skimr")
library("dplyr")
library("RcmdrMisc")
library("haven")
library("sjPlot")
library(ggplot2)
``` 



# Blood BM
## Cardiovascular Indicators (bp)
```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
bp <- read_sas("data/evw0173.sas7bdat")
nrow(bp) # 2621
length(unique(bp$publicID)) # 674
skim(bp)
```
There are a lot of missing data points for diastol3 and systol3. They take three measurements just to make sure they are somewhat consistent. We can average all three, and if any one is missing, just average whatever numbers we have. If all three are missing, then it's a NaN

```{r}
bp$syst_avg <- rowMeans(subset(bp, select = c(`systol1`, `systol2`, `systol3`)), na.rm = TRUE)
bp$dia_avg <- rowMeans(subset(bp, select = c(`diastol1`, `diastol2`, `diastol3`)), na.rm = TRUE)

skim(bp)
```


# Only get the average systol and diastol and other relevent info
```{r}
bp_avg <- subset(bp, select = c(`publicID`, `week`, `syst_avg`, `dia_avg`))

length(bp_avg$publicID) #2570
length(unique(bp_avg$publicID)) # 673
table(bp_avg$week)
```

```{r}
skim(bp_avg)

bp_avg %>% group_by(week) %>% summarize(sum_na_syst = sum(is.na(syst_avg)),
                                        sum_na_dia = sum(is.na(dia_avg)))
```
Not many missing


# LBW0070 Routine Chemistries
```{r}
mm <- read_sas("data/lbw0070.sas7bdat")

length(unique(mm$publicID))

skim(mm)
```

# Get only values we are interested in
```{r}
mm <- subset(mm, select = -c(`bunlon`, `bungrd`, `glucval`:`pyru1val`, `othchem`:`othage3`, `visitobt`))
skim(mm)

```
sgptval -  ALT/SGPT value
sgptlon - ALT/SGPT ULN (upper limit normal)
sgptgrd - ALT/SGPT grade

sgotval- AST/SGOT value
sgotlon - AST/SGOT ULN
sgotgrd - AST/SGOT grade
NOTES: We may not need ULN. Grade is telling is how bad this value is ordinally (See data dictionary).

bunval - BUN value mg/dl

ckval - Creatine kinase value
cklon - Creatine kinase ULN
ckgrd - Creatine kinase grade

lipval - Lipase value
liplon - Lipase ULN
lipgrd - Lipase grade 

cretval - Serum Creatine kinase value
cretlon - Serum Creatine kinase ULN
cretgrd - Serum Creatine kinase grade

To Take out:
enrsmart
amppos
chemres

ftcval - fasting total cholesterol value
ftcgrd - fasting total cholesterol grade

ftgval - fasting triglycerides value
ftggrd - fasting triglycerides grade

fldlval - Fasting LDL value mg/dl
fldlgrd - Fasting LDL grade

fhdlval - Fasting HDL value mg/dl
fhdlgrd - Fasting HDL grade

everything after ochem is discontinued..


# Find duplicated rows and merge them
NB: If there are multiple values for the same measure, average them. However, there are very few
cases of this
```{r}
mm_dup <- mm[duplicated(mm[,c(1,3)]) | duplicated(mm[,c(1,3)], fromLast = T), ]
mm_not_dup <- mm[!(duplicated(mm[,c(1,3)]) | duplicated(mm[,c(1,3)], fromLast = T)), ]
mm_dup_merged <- mm_dup %>% group_by(publicID, week) %>% 
  summarise(across(everything(), mean, na.rm = TRUE), .groups = 'drop')

```

# Merge fixed duplicated and regular dataframe
```{r}
mm_dup_merged$study <- as.character(mm_dup_merged$study)

mm_fix <- rbind(mm_dup_merged, mm_not_dup) 
nrow(mm_fix) #2051 This is close to other dataset sizes
length(unique(mm_fix$publicID))
```


```{r}
table(mm_fix$week)

mm_fix_3 <- subset(mm_fix, week %in% c(0, 48, 96))
table(mm_fix_3$week)

length(mm_fix_3$publicID) #1321
length(unique(mm_fix_3$publicID)) # 560

skim(mm_fix_3)
```


# NaN's in each column
```{r}
na_count <-sapply(mm_fix_3, function(y) sum(length(which(is.na(y)))))

# Percent of NA's in each column
na_count/1059
```
Nothing has more less than 50% complete


# LBW0073 Routine Hematologies
```{r}
hem <- read_sas("data/lbw0073.sas7bdat")
length(unique(hem$publicID))

```


# Get only columns we are interested in
```{r}
hem <- subset(hem, select = -c(`atlymph`, `othval1`:`oth_gd2`, `othage1`:`othage2`))
skim(hem)
```

```{r}
table(hem$week)

# length(hem[complete.cases(hem),]$publicID) #1842
# length(unique(hem[complete.cases(hem),]$publicID)) # 447
# comp_hem <- hem[complete.cases(hem),]
```

Only 44% of alympcnt (Absolute lymphocyte count (mm^3)) and alympgrd (grade) are complete at baseline. We can take this out.


```{r}
hem_3 <- subset(hem, week %in% c(0, 48, 96))
table(hem_3$week)

hem_3 <- hem_3 %>% group_by(publicID) %>% mutate(total = n())

length(unique(hem_3$publicID)) # 450

hem_3 <- subset(hem_3, select = -c(`hemadone`, `othhemob`, `formver`, `oth1`, `oth2`, `total`))

```

# NaN's in each column
```{r}
na_count_hem <-sapply(hem_3, function(y) sum(length(which(is.na(y)))))

# Percent of NA's in each column
na_count_hem/1298
```
Nothing here has more less than 50% complete.


# Height, weight, body measurements (i.e., waist, hip) EVW0169
```{r}
gro <- read_sas("data/evw0169.sas7bdat")

skim(gro)

gro <- subset(gro, select = c(`publicID`, `week`, `height1`, `height2`, `height3`,
                              `weight1`, `weight2`, `weight3`,
                              `mwc1`, `mwc2`, `mwc3`))
gro$weight_avg <- rowMeans(subset(gro, select = c(`weight1`, `weight2`, `weight3`)), na.rm = TRUE)
gro$height_avg <- rowMeans(subset(gro, select = c(`height1`, `height2`, `height3`)), na.rm = TRUE)
gro$mwaist_avg <- rowMeans(subset(gro, select = c(`mwc1`, `mwc2`, `mwc3`)), na.rm = TRUE)
gro$bmi <- gro$weight_avg / ((gro$height_avg)/100)^2

gro_avg <- subset(gro, select = c(`publicID`, `week`, `weight_avg`, `height_avg`, `bmi`, `mwaist_avg`))
```

The formula is BMI = kg/m2 where kg is a person’s weight in kilograms and m2 is their height in metres squared.
```{r}
table(gro_avg$week)

gro_avg_3 <- subset(gro_avg, week %in% c(0, 48, 96))
gro_avg_3 <- gro_avg_3 %>% group_by(publicID) %>% mutate(total = n())

length(unique(gro_avg_3$publicID)) # 674
gro_avg_3 <- gro_avg_3 %>% select(-c(total))
```
# See how many NaNs in each column
```{r}
na_count_gro <-sapply(gro_avg_3, function(y) sum(length(which(is.na(y)))))

# Percent of NA's in each column
na_count_gro/1827
```
Nothing has more than 50% missing

# Get HIV status DMW0080
```{r}
dem <- read_sas("data/dmw0080.sas7bdat")
table(dem$hivinf)

dem <- subset(dem, select = c(`publicID`, `week`, `hivinf`))
dem[dem$publicID == 95804,]
dem_3 <- dem %>% group_by(publicID) %>% summarize(hivinf1 = mean(hivinf, na.rm = TRUE))
table(dem_3$hivinf1)
```


# Merging
```{r}
nrow(bp_avg) # 2621
nrow(mm_fix_3) # 1321
m1 <- merge(bp_avg, mm_fix_3, by = c("publicID", "week"))
nrow(m1) # 1316

nrow(hem_3) # 1298
m2 <- merge(m1, hem_3, by = c("publicID", "week"))
nrow(m2) # 853

nrow(gro_avg_3) # 1827
m3 <- merge(m2, gro_avg_3, by = c("publicID", "week"))
nrow(m3) # 864

nrow(dem_3) # 669
m4 <- merge(m3, dem_3, by = c("publicID"))
nrow(m4) # 864

length(unique(m4$publicID))
```

# Overall characteristics
```{r}
# Number of NA's by row (observation for participant at time point x)
na_count_final_rows <- apply(X = is.na(m4), MARGIN = 1, FUN = sum)

# sort(na_count_final_rows/(ncol(m4) - 5), decreasing = T)[0:30]
# There are some observations where most of the datapoints are missing across columns

less_50 <- which(na_count_final_rows/(ncol(m4) - 5) >= 0.50)
length(less_50)
# Indices of those with more than 50% of BM's missing

# Get rid of those observations
m4 <- m4[-less_50,]
m4 <- m4 %>% select(-c(study.x, hivinf1, chemobt))

length(unique(m4$publicID))

# Number of NA's in each column
na_count_final_cols <-sapply(m4, function(y) sum(length(which(is.na(y)))))

sort(na_count_final_cols/nrow(m4))
```
No column has more than 50% missing


# Overall Characteristics By Week
```{r}
m4_0 <- m4[m4$week == 0,]
m4_48 <- m4[m4$week == 48,]
m4_96 <- m4[m4$week == 96,]

# Returns list with two objects
# 1. indices of df (observations) where row has less than 50% data available
# 2. Percentage complete of each column.
na_count_rows_cols <- function(df) {
  
  # Number of NA's by row (observation for participant at time point x)
  count_rows <- apply(X = is.na(df), MARGIN = 1, FUN = sum)
  less_50 <- which(count_rows/(ncol(df) - 4) >= 0.50)
  
  # Number of NA's in each column
  count_cols <- sapply(df, function(y) sum(length(which(is.na(y)))))
  na_cols_perc <- sort((count_cols / nrow(df)), decreasing = T)
  
  return(list(less_50, na_cols_perc))
}

# Run function by week
na_count_rows_cols(m4_0)
na_count_rows_cols(m4_48)
na_count_rows_cols(m4_96)

```
NB: alympcnt (Lymphocyte count) and alympgrd (Lymphocyte Grade) only have about 32% completed at baseline
There are a good amount of individual observations missing at 48 weeks.


# Histograms for BM's 
```{r}
m4_cont <- m4 %>% select(c(syst_avg, dia_avg, sgptval, sgotval, bunval, ckval, lipval, cretval,
                           ftcval, ftgval, fldlval, fhdlval, hemocv, hema, rbc, mcv, wbccv, neutcv,
                           anccv, lymph, mono, eosin, baso, platcv, specage, weight_avg, height_avg,
                           bmi, mwaist_avg))


m4_list_cont <- colnames(m4_cont)

for (coln in m4_list_cont) {
  print(
    ggplot(data = m4, aes_string(coln)) + 
  geom_histogram(bins = 30) + 
  facet_grid(. ~ week) +
  labs(title = paste(coln, "by Week"),
       x = coln,
       y = "Frequency") + 
  theme(plot.title = element_text(hjust = 0.5))
  )
}


```

# Make Tables
```{r}
m4_0_col_perc <- data.frame(na_count_rows_cols(m4_0)[2])
m4_48_col_perc <- data.frame(na_count_rows_cols(m4_48)[2])
m4_96_col_perc <- data.frame(na_count_rows_cols(m4_96)[2])

# rename columns
m4_0_col_perc <-  cbind(Measure = rownames(m4_0_col_perc), m4_0_col_perc)
rownames(m4_0_col_perc) <- 1:nrow(m4_0_col_perc)
colnames(m4_0_col_perc)[2] <- "Prop. Complete (Week 0)"

# cbind 48 and 96 to the df
colnames(m4_48_col_perc)[1] <- "Prop. Complete (Week 48)"
colnames(m4_96_col_perc)[1] <- "Prop. Complete (Week 96)"

bmt1 <- cbind(m4_0_col_perc, m4_48_col_perc)
bm_cols_table <- cbind(bmt1, m4_96_col_perc)

# Export to Doc file
tab_df(bm_cols_table, file = "test.doc")

```


# Final BM's
```{r}
m4_fin <- m4 %>% select(-c(sgptlon, sgptgrd, sgotlon, sgotgrd, cklon, ckgrd, liplon, lipgrd, cretlon,
                          cretgrd, enrsmart, amppos, chemres, ftcgrd, ftggrd, fldlgrd, fhdlgrd, 
                          study.y, hemo_gd, anc_gd, plat_gd, alympgrd, plat_gd, alympgrd, wbcgrd))

skim(m4_fin)
length(unique(m4_fin$publicID)) # 286
```

# Stargazer
```{r}
#install.packages("stargazer")
library(stargazer)
```

# Tables for participants
```{r}
dem <- read_sas("/Users/hansoochang/Drexel/PHACS:AMP/data/master.sas7bdat")
dem <- dem %>% select(publicID, entryage, sex, prace, pethnc)

table(dem$prace)

m4_dem <- merge(m4_fin, dem, by = c("publicID")) %>% 
  select(publicID, entryage, week, sex, prace, pethnc)

m4_dem %>% group_by(sex, week) %>% summarise(count = n(), )

# Change variable names race
m4_dem$prace[m4_dem$prace == 1001] <- "American Indian/Alaska Native"
m4_dem$prace[m4_dem$prace == 1002] <- "Asian"
m4_dem$prace[m4_dem$prace == 1003] <- "Native Hawiaiian or other Pacific Islander"
m4_dem$prace[m4_dem$prace == 1004] <- "Black or African American"
m4_dem$prace[m4_dem$prace == 1005] <- "White"
m4_dem$prace[m4_dem$prace == 1006] <- "American Indian"
m4_dem$prace[m4_dem$prace == 1007] <- "Alaskan Native"
m4_dem$prace[m4_dem$prace == 8000] <- "Other"
m4_dem$prace[m4_dem$prace == 9000] <- "More than One Race"
m4_dem$prace <- as.factor(m4_dem$prace)

# Change variable names in 
m4_dem$sex[m4_dem$sex == 1] <- "M"
m4_dem$sex[m4_dem$sex == 2] <- "F"
m4_dem$sex <- as.factor(m4_dem$sex)

m4_dem$entryage <- as.numeric(m4_dem$entryage)

m4_dem %>% filter(week == 0) %>% count(sex)
m4_dem %>% filter(week == 0) %>% summarise(mean = mean(entryage))
nrow(m4_dem %>% filter(week == 0))

```

## Demograhic info
```{r}
# Count sex
res <- m4_dem %>% group_by(publicID) %>%
  summarise(male_count = sum(sex == "M", na.rm = TRUE),
            female_count = sum(sex == "F", na.rm = TRUE)) 

count_males <- sum(res$male_count != 0)
count_males


# Count Race
res_race <- m4_dem %>% group_by(publicID) %>%
  summarise(AIAN_count = sum(prace == "American Indian/Alaska Native", na.rm = TRUE),
            Asian_count = sum(prace == "Asian", na.rm = TRUE),
            NHPI_count = sum(prace == "Native Hawiaiian or other Pacific Islander", na.rm = TRUE),
            Black_count = sum(prace == "Black or African American", na.rm = TRUE),
            White_count = sum(prace == "White", na.rm = TRUE),
            AI_count = sum(prace == "American Indian", na.rm = TRUE),
            AN_count = sum(prace == "Alaskan Native", na.rm = TRUE),
            Other_count = sum(prace == "Other", na.rm = TRUE),
            Multi_count = sum(prace == "More than One Race", na.rm = TRUE))

count_race <- colSums(res_race != 0)
count_race
```


# Tables for BM's
```{r}
m4_bms <- m4_fin %>% select(-c(publicID))
  
m4_bms %>% group_by(week) %>% summarize()

skim(m4_bms)
  
```


# Export
Files are demographic info, biomarker info 
```{r}
write.csv(m4_dem, "/Users/hansoochang/Drexel/PHACS:AMP/Exported_Data/m4_dem.csv")
write.csv(m4_fin, "/Users/hansoochang/Drexel/PHACS:AMP/Exported_Data/m4_fin.csv")


```



NB: We can add fasting glucose via  LBW0071

### Cardiovascular Indicators (bp)
Average Blood Pressure

### Metabolic Markers (total cholesterol, fasting glucose) ###
ftcval - fasting total cholesterol value 
ftcgrd - fasting total cholesterol grade

fldlval - Fasting LDL value mg/dl
fldlgrd - Fasting LDL grade

fhdlval - Fasting HDL value mg/dl
fhdlgrd - Fasting HDL grade


### Markers of inflammation and infection (e.g., cytomegalovirus, C-reactive protein, albumin, IL-6) ###
lipval - Lipase value
liplon - Lipase ULN
lipgrd - Lipase grade 

(From tology)
anccv - Abs count neutrophil value (mm^3)
platcv - Platelets value (mm^3)
hemocv - Hemoglobin value (g/dl)
rbc - RBC value (million/ul)
mcv - MCV value (microns)
wbccv - WBC value (mm^3)
lymphocyte count


### Markers of organ functioning (serum creatinine, peak flow, BUN, alkaline phosphatase) ###
sgptval -  ALT/SGPT value
sgptlon - ALT/SGPT ULN (upper limit normal)
sgptgrd - ALT/SGPT grade

sgotval- AST/SGOT value
sgotlon - AST/SGOT ULN
sgotgrd - AST/SGOT grade
NOTES: We may not need ULN. Grade is telling is how bad this value is ordinally (See data dictionary).

cretval - Serum Creatine kinase value
cretlon - Serum Creatine kinase ULN
cretgrd - Serum Creatine kinase grade

bunval - BUN value mg/dl

## i.e., creatinine, blood urea nitrogen, AST, ALT, lipids, fasting glucose and insulin, fasting lipids and renal and cardiac biomarkers

# Height, weight, body measurements (i.e., waist, hip)

# Urine BM's SPW0396
```{r}
# uri <- read_sas("data/spw0396.sas7bdat")
# uri
```

### Cardiovascular Indicators (bp)
Average Blood Pressure

### Metabolic Markers (total cholesterol, fasting glucose) ###
Values from SPW0396, but categorical..

### Markers of inflammation and infection (e.g., cytomegalovirus, C-reactive protein, albumin, IL-6) ###
Values from SPW0396, but categorical..

### Markers of organ functioning (serum creatinine, peak flow, BUN, alkaline phosphatase) ###
Values from SPW0396, but categorical..

# Height, weight, body measurements (i.e., waist, hip)
Height
Weight
BMI

















