---
title: "PoA_V3"
output:
  html_document:
    toc: true
    theme: united
date: "2023-08-29"
---


This file calculates the Pace of Aging for all three methods. *** Version 3 with biomarkers chosen by Step AIC

Read in:

  * ivs_stand_flipped_v3.csv (from Biomarker_Cleaning.Rmd)
  * hxw0101.sas7bdat (Viral Load)

Output: 

  * Dataframes with Pace of Aging for each method


# Libraries
```{r,  echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
library("skimr")
library("dplyr")
library("RcmdrMisc")
library("haven")
library("sjPlot")
library(ggplot2)
library(purrr)
library(tidyr)
library(ggrepel)
library(lme4)
library(LDATS)
library(dplyr)
library(BioAge)
library(Metrics)
library(ggpubr)
library(broom)
library(tibble)
```

# Read in Dataframe
```{r}

# Biomarkers
ivs_stand <- read.csv(paste0(root, "/data/ivs_stand_flipped_v3.csv"))

ivs_stand %>% filter(week == 0) %>% summarise(
  min_age = min(specage.y, na.rm = TRUE),
  max_age = max(specage.y, na.rm = TRUE))

ivs_stand <- ivs_stand %>% select(-specage.x)

# Viral Load
hxw <- read_sas(paste0 (root, "/data/hxw0101.sas7bdat"))

v_load <- hxw %>% select(week, visitage, rnaage, rnaobt, rna, rnabeage, rnaafage)

```


# Get Random Intercept for each biomarker for each Participant
## Our Method
This chunk also runs an anova to test whether including a random slope and intercept improves the model
```{r, warning=FALSE, error=FALSE, message=FALSE}

biomarkers <- ivs_stand %>%
  select(sgptval:mwaist_avg) %>%
  names()


# Create an empty dataframe to store the random effects
random_intercept_df <- data.frame(id = unique(ivs_stand$publicID))
random_slope_df <- data.frame(id = unique(ivs_stand$publicID))
fixed_slope_df <- data.frame(biomarker = NA)
anova_results <- data.frame(biomarker = NA, p_value = NA)

# This for loop extracts the fixed and random slope and intercept for each biomarker
for (bm in biomarkers) {
  print(bm)
  # Create the formula dynamically
  formula <- as.formula(paste(bm, "~ 1 + scale_base_age*week_scale + (1 + week_scale | publicID)"))
  
  # Fit the lmer model
  model <- lmer(formula, data = ivs_stand)
  
  # Extract the random effects of the intercept
  random_intercept <- as.data.frame(ranef(model)$publicID[,1])
  random_slope <- as.data.frame(ranef(model)$publicID[,2])
  fixed_slope <- as.data.frame(fixef(model)[2])
  

  # Add the random effects to the dataframe
  random_intercept_df <- merge(random_intercept_df, random_intercept, by = 0, all = T)
  names(random_intercept_df)[length(names(random_intercept_df))] <- bm 
  
  random_slope_df <- merge(random_slope_df, random_slope, by = 0, all = T)
  names(random_slope_df)[length(names(random_slope_df))] <- bm 
  
  fixed_slope_df[,bm] <- fixed_slope
  
  # Fit the lmer model without random effects
  model_no_re <- lm(paste(bm, "~ 1 + scale_base_age*week_scale"), data = ivs_stand)
  
  # Run ANOVA test to compare models
  anova_result <- anova(model, model_no_re)
  p_value <- anova_result[["Pr(>Chisq)"]][2]  # Extract the p-value for the random effects
  
  anova_results <- rbind(anova_results, data.frame(biomarker = bm, p_value = p_value))
}

# Filter significant results for ANOVA
significant_results <- anova_results %>% filter(p_value < 0.05)
significant_results

# Clean up the dataframe
random_intercept_df <- random_intercept_df %>%
  select(-matches("Row.names"))

random_slope_df <- random_slope_df %>%
  select(-matches("Row.names"))

```

# Run softmax function for the intercepts
```{r}

# Take out ID column
df_softmax <- random_intercept_df %>% select(-id)

# Make all NA's into 0's
df_softmax <- df_softmax %>% mutate_all(~replace(., is.na(.), 0))

# Apply softmax function across rows
df_softmax <- apply(df_softmax, 1, softmax)
 
df_softmax <- cbind(random_intercept_df$id, as.data.frame(t(df_softmax)))

skim(df_softmax)
# make sure all the rows add up to one
rowSums(df_softmax[,-1])

```

# Calculate Belsky 2015
```{r, warning=FALSE, error=FALSE, message=FALSE}
biomarkers <- ivs_stand %>%
  select(sgptval:mwaist_avg) %>%
  names()

# Create an empty dataframe to store the random effects
random_slope_belsky <- data.frame(publicID = unique(ivs_stand$publicID), biomarker = NA)


# This for loop extracts the random slope for each biomarker
for (bm in biomarkers) {
  # Create the formula dynamically
  formula <- as.formula(paste(bm, "~ 1 + scale_base_age*week_scale + (1 + week_scale | publicID)"))
  
  # Fit the lmer model
  model <- lmer(formula, data = ivs_stand)
  
  # Extract the random effects of the slope
  random_slope <- as.data.frame(ranef(model)$publicID[,2])
  

  # Add the random effects to the dataframe
  random_slope_belsky <- merge(random_slope_belsky, random_slope, by = 0, all = T)
  
  names(random_slope_belsky)[length(names(random_slope_belsky))] <- bm 

}

# Clean up the dataframe
random_slope_belsky <- random_slope_belsky %>%
  select(-matches("Row.names"))
random_slope_belsky <- random_slope_belsky %>% select(-biomarker)

# Print the resulting dataframe
skim(random_slope_belsky)

```


# KDM
## Split and train on baseline training
Not enough participants to split by gender
```{r}
set.seed(18)

# Create age column and clean # 
ivs_stand$age <- ivs_stand$specage.y
ivs_stand <- ivs_stand %>% select(-contains("X."))

# ### Males ### 
# ivs_stand_m <- ivs_stand %>% filter(sex == "M") 
# ivs_stand_m_sample <- sample(c(TRUE, FALSE), nrow(ivs_stand_m), 
#                              replace=TRUE, prob=c(0.5,0.5))
# male_train <- ivs_stand_m[ivs_stand_m_sample, ]
# male_test <- ivs_stand_m[!ivs_stand_m_sample, ]
# 
# ### Females ###
# ivs_stand_f <- ivs_stand %>% filter(sex == "F") 
# ivs_stand_f_sample <- sample(c(TRUE, FALSE), nrow(ivs_stand_f), 
#                              replace=TRUE, prob=c(0.5,0.5))
# female_train <- ivs_stand_f[ivs_stand_f_sample, ]
# female_test <- ivs_stand_f[!ivs_stand_f_sample, ]

# Split into time points
ivs_stand_base <- ivs_stand %>% filter(week == 0) 
ivs_stand_48 <- ivs_stand %>% filter(week == 48) 
ivs_stand_96 <- ivs_stand %>% filter(week == 96) 
```

# Remove Outliers in baseline training
## Graph Distribution
```{r}
biomarkers <- ivs_stand_base %>% select(sgptval:mwaist_avg) %>% names()

for (bm in biomarkers) {
  hist(ivs_stand_base[,bm], main = paste("Histogram of", bm))
}
```

## remove any outliers
```{r}
ivs_stand_base <- ivs_stand_base %>%
  mutate(across(biomarkers, ~ ifelse(abs(.) > 4, NA, .)))
```

```{r}

# Train Baseline
ivs_stand_base_sample <- sample(c(TRUE, FALSE), nrow(ivs_stand_base), 
                             replace=TRUE, prob=c(0.5,0.5))
ivs_stand_base_train <- ivs_stand_base[ivs_stand_base_sample, ]
ivs_stand_base_test <- ivs_stand_base[!ivs_stand_base_sample, ]

```


## Train KDM
```{r}
markers <- ivs_stand_base %>%
  select(cretval:mwaist_avg) %>%
  names()

kdm_train_base <- kdm_calc(ivs_stand_base_train, biomarkers = markers)

kdm_train_base_comp <- kdm_train_base$data %>% filter(!is.na(kdm))

```

# KDM Correlation tests between kdm_age and chronological age
```{r}
cor.test(as.numeric(kdm_train_base_comp$kdm), as.numeric(kdm_train_base_comp$age))
rmse(as.numeric(kdm_train_base_comp$kdm), as.numeric(kdm_train_base_comp$age))
mae(as.numeric(kdm_train_base_comp$kdm), as.numeric(kdm_train_base_comp$age))

```

## Project to test
```{r}
kdm_test_base <- kdm_calc(ivs_stand_base_test, biomarkers = markers, fit = kdm_train_base$fit, s_ba2 = 
                        kdm_train_base$fit$s_ba2)

kdm_test_base_comp <- kdm_test_base$data %>% filter(!is.na(kdm))
```

## Correlation tests between kdm_age (both genders) and chronological age
```{r}
cor.test(as.numeric(kdm_test_base_comp$kdm), as.numeric(kdm_test_base_comp$age))
rmse(as.numeric(kdm_test_base_comp$kdm), as.numeric(kdm_test_base_comp$age))
mae(as.numeric(kdm_test_base_comp$kdm), as.numeric(kdm_test_base_comp$age))

kdm_test_base_comp$gender <- kdm_test_base_comp$sex
#plot age vs bioage
plot_ba(kdm_test_base_comp, "kdm", "KDM\nBiological Age")

#plot age vs bioage
plot_ba(kdm_test_base_comp, "kdm_advance", "KDM\nBiological Age")

```

# Project KDM to next time points
```{r}
# 48 weeks
kdm_project_48 <- kdm_calc(ivs_stand_48, biomarkers = markers, fit = kdm_train_base$fit, s_ba2 = 
                        kdm_train_base$fit$s_ba2)

kdm_project_48_comp <- kdm_project_48$data %>% filter(!is.na(kdm))

# 96 weeks
kdm_project_96 <- kdm_calc(ivs_stand_96, biomarkers = markers, fit = kdm_train_base$fit, s_ba2 = 
                        kdm_train_base$fit$s_ba2)

kdm_project_96_comp <- kdm_project_96$data %>% filter(!is.na(kdm))



```

# Correlation and Metrics for Future Projections
```{r}
# 48 weeks 
nrow(kdm_project_48_comp)
cor.test(as.numeric(kdm_project_48_comp$kdm), as.numeric(kdm_project_48_comp$age))
rmse(as.numeric(kdm_project_48_comp$kdm), as.numeric(kdm_project_48_comp$age))
mae(as.numeric(kdm_project_48_comp$kdm), as.numeric(kdm_project_48_comp$age))

# 96 weeks 
nrow(kdm_project_96_comp)
cor.test(as.numeric(kdm_project_96_comp$kdm), as.numeric(kdm_project_96_comp$age))
rmse(as.numeric(kdm_project_96_comp$kdm), as.numeric(kdm_project_96_comp$age))
mae(as.numeric(kdm_project_96_comp$kdm), as.numeric(kdm_project_96_comp$age))
```


# Look at KDM Longitudinally
```{r}
kdm_project_48_comp$gender <- kdm_project_48_comp$sex
kdm_project_96_comp$gender <- kdm_project_96_comp$sex

kdm_all_comp <- rbind(kdm_project_96_comp, rbind(kdm_test_base_comp, kdm_project_48_comp))

# Add a column to signify KDM when comparing with other PoA methods.
kdm_all_comp$group <- "kdm"

kdm_all_comp_3weeks <- kdm_all_comp %>%
  group_by(publicID) %>%
  filter(n() == 3) %>%
  ungroup()

ggplot(kdm_all_comp_3weeks, aes(x = age, y = kdm, group = publicID)) +
  geom_point(aes(color = gender)) + 
  geom_line(aes(color = gender)) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  labs(x = "Age", y = "KDM") +
  theme_minimal()

hist(kdm_all_comp$kdm_advance)
```


# KDM Pace of Aging
```{r}
# All participants with at least 2 time points.
kdm_all_comp_2weeks <- kdm_all_comp %>%
  group_by(publicID) %>%
  filter(n() >= 2) %>%
  ungroup()

# Get individual models and slopes
# individual_models <- kdm_all_comp_2weeks %>%
#   group_by(publicID) %>%
#   do(beta = summary(lm(kdm ~ week_scale * scale_base_age, data = .))$coefficients["week_scale",1],
#      p_val = summary(lm(kdm ~ week_scale * scale_base_age, data = .))$coefficients["week_scale",4])
# 
# individual_models <- unnest(individual_models)


# Fit a linear mixed-effects model
kdm_mixed_model <- lmer(kdm ~ week_scale * scale_base_age + (week_scale | publicID), 
                    data = kdm_all_comp_2weeks)

# Extract random effects
random_effects <- ranef(kdm_mixed_model)

# Make into dataframe
kdm_re <- random_effects$publicID %>%
  rownames_to_column(var = "publicID")
kdm_re$publicID <- as.integer(kdm_re$publicID)


# Rename column and add one
kdm_re <- kdm_re %>%
  rename(poa = week_scale)

# Add 1
kdm_re$poa <- kdm_re$poa + 1
```

# Merge KDM dataframe with KDM PoA
```{r}

kdm_poa <- right_join(kdm_all_comp %>% select(publicID, sgptval:mwaist_avg, group),
      kdm_re %>% select(publicID, poa),
      by = "publicID")
```

# Distribution of KDM PoA
```{r}
hist(kdm_poa$poa)
sd(kdm_poa$poa)

```


# Predict Pace of Biological Aging for Current Participants
## Our method random slope and random intercept
Pace of BAi = sumj = 1(Wji * m1ji)
```{r}
# Multiply Matrices
combo <- random_slope_df[,-1] * df_softmax[, -1]
# Replace NA values with column averages
for(i in 1:ncol(combo)){ 
  combo[is.na(combo[,i]), i] <- mean(combo[,i], na.rm = TRUE)
}

# Calculated pace of aging (sum of rows)
combo$poa <- round(rowSums(combo), digits = 5)

# add public id to dataframe
combo$publicID <- random_slope_df$id

combo$poa <- combo$poa + 1

hist(combo$poa)

sd(combo$poa)
combo$poa
```


# Belsky 2015 Prediction
Pace of BAi = sumj = 1(m1ji / number of biomarkers)
```{r}
# Select the range of columns you want to divide by the length
cols_to_divide <- random_slope_belsky %>% select(cretval:mwaist_avg) %>% names()

random_slope_belsky <- random_slope_belsky %>%
  mutate(across(all_of(cols_to_divide), ~ . / length(cols_to_divide)))


# Replace NA values with column averages
for(i in 1:ncol(random_slope_belsky)){ 
  random_slope_belsky[is.na(random_slope_belsky[,i]), i] <- mean(random_slope_belsky[,i], na.rm = TRUE)
}


# Calculated pace of aging (sum of rows)
random_slope_belsky$poa <- round(rowSums(random_slope_belsky %>% select(cols_to_divide)), digits = 5)
random_slope_belsky$poa <- random_slope_belsky$poa + 1
hist(random_slope_belsky$poa)


```


# Export PoA Prediction
```{r}
combo$group <- "Chang"
random_slope_belsky$group <- "Belsky"

write.csv(combo, paste0(root, "Exported_Data/chang.csv"))
write.csv(random_slope_belsky, paste0(root, "Exported_Data/belsky.csv"))
write.csv(kdm_poa, paste0(root, "Exported_Data/kdm.csv"))

# KDM by time point (Not PoA)
write.csv(kdm_all_comp, paste0(root, "Exported_Data/kdm_all_comp.csv"))


length(unique(combo$publicID))
length(unique(random_slope_belsky$publicID))
length(unique(kdm_poa$publicID))

mean(combo$poa)
sd(combo$poa)
mean(belsky$poa)
sd(belsky$poa)
mean(kdm_all_comp$kdm_advance)
sd(kdm_all_comp$kdm_advance)

```






















